 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter*{LIST OF PUBLICATIONS AND AUTHOR CONTRIBUTIONS}
%\label{sec:pubs}
\addcontentsline{toc}{chapter}{LIST OF PUBLICATIONS AND AUTHOR CONTRIBUTIONS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the course of satisfying the University of Arizona Department of Physics's requirements for a 
Ph.D. doctoral dissertation, I listed the following publications and point out my contribution to each work which is described under each item.
\begin{itemize}
 \item ''A Short Survey of Matter-Antimatter Evolution in the Primordial 
Universe'' by~[\cite{Rafelski:2023emw}] is a 50 page long review with many novel results 
describing the role of antimatter in the early universe. I collaborate with 
Andrew Steinmetz, Jeremiah Birrell, and Johann Rafelski the document creation, providing the numerical results/figures for the paper to help creating one coherent presentation. I acknowledge the help and consultation of Andrew Steinmetz, Jeremiah Birrell, and Johann Rafelski in research and computation. 

\item ''Matter-antimatter origin of cosmic magnetism'' 
by~[\cite{Steinmetz:2023nsc}] proposes a model of para-magnetization driven by the large 
matter-antimatter (electron-positron) content of the early universe. I collaborate with Andrew Steinmetz to carry out the computation
, contribute key results and five 
technical figures for the paper. I acknowledge the help and consultation of Andrew Steinmetz, Martin Formanek, and Johann Rafelski in research and computation. 

\item ''Electron-positron plasma in BBN: Damped-dynamic screening'' by~[\cite{Grayson:2023flr}] use the linear response theory adapt by C. Grayson to describe the inter nuclear potential in electron/positron plasma. My contribution is the computation of the chemical potential and plasma damping rate which are important properties to  implement into relativistic Boltzmann equation and linear response theory. I acknowledge the help and consultation of Christopher Grayson, Martin Formanek, and Johann Rafelski in research and computation.

\item ''Cosmological strangeness abundance'' by~[\cite{Yang:2021bko}] investigates the strange particle composition of the expanding early Universe and examine their freeze-out temperatures. I performed all computation, writing, and 
figure making for the paper. I acknowledge the help and consultation of Johann Rafelski in 
research, writing and editing.

\item ''The muon abundance in the primordial Universe'' by~[\cite{Rafelski:2021aey}] is a study of the muon abundance and its persistence temperature in early Universe. I performed all 
computation and writing in preparation of the first draft and approved the final draft before 
submission. I acknowledge the help and consultation of Johann Rafelski in research, writing and editing.

\item ''Reactions Governing Strangeness Abundance in Primordial Universe'' by~[\cite{Rafelski:2020ajx}] is a conference proceeding paper for
the 19th International Conference on Strangeness in Quark Matter (SQM 2021). It summarize our earlier work in strangeness reactions. I performed all 
computation and writing in preparation of the first draft and approved the final draft before 
submission. I acknowledge the help and consultation of Johann Rafelski in research, writing and editing.

\item''Possibility of bottom-catalyzed matter genesis near to primordial QGP hadronization'' by~[\cite{Yang:2020nne}] is our fist study of the bottom flavor abundance and show the  nonequilibrium behavior near to QGP hadronization. I performed all computation and writing in preparation of the paper. I acknowledge the help and consultation of Johann Rafelski in research, writing and editing.

\item ''Lepton Number and Expansion of the Universe'' by~[\cite{Yang:2018oqg}] proposes a model of large lepton asymmetry and explore how this large cosmological lepton yield
relates to the effective number of (Dirac) neutrinos. I performed all computation and writing in preparation of the paper. I acknowledge the help and consultation of Johann Rafelski in research, writing and editing.

\item''Temperature Dependence of the Neutron Lifespan'' by~[\cite{Yang:2018qrr}] is a study of neutron lifespan in plasma with Fermi-blocking from electron and neutrino. I performed all computation and writing in preparation of the paper. I acknowledge the help and consultation of Johann Rafelski in research, writing and editing.

\item ''Strong fields and neutral particle magnetic moment dynamics'' by~[\cite{Formanek:2017mbv}] is an overview of our research group's efforts in studying neutral particle dynamics in electromagnetic fields. My contribution is on the neutrino section. I consulted and helped lead author Martin Formanek, and co-authors Stefan Evans, Andrew Steinmetz, and Johann Rafelski in 
editing and revising the manuscript.

\item "Relic Neutrino Freeze-out: Dependence on Natural Constants" by~[\cite{Birrell:2014uka}] is a study of neutrino freeze-out temperature as a function of standard model parameter and its application on the effective number of (Dirac) neutrinos. My contribution is the calculation of all the neutrino-matter weak interaction matrix elements required for the Boltzmann code. I acknowledge the help and consultation of Jeremiah Birrell and Johann Rafelski in research and computation.

\item ''Fugacity and Reheating of Primordial Neutrinos'' by~[\cite{Birrell:2013gpa}] is as study of neutrino fugacity as a funciton of neutrin kinetic freeze-out temperature. My contribution is the calculation of neutrino interaction matrix  elements and helping the evaluation of  neutrino relaxation time. I acknowledge the help and consultation of Jeremiah Birrell, Johann Rafelski, and Pisin Chen in research and computation.

\item''Relic neutrinos: Physically consistent treatment of effective number of neutrinos and neutrino mass'' by~[\cite{Birrell:2012gg}] is a model independent study of the neutrino momentum distribution at freeze-out, treating the freeze-out temperature as a free parameter. I collaborate with Jeremiah Birrell, Johann Rafelski, and Pisin Chen the document creation. I acknowledge the help and consultation of Jeremiah Birrell, Johann Rafelski, and Pisin Chen in research. 

\end{itemize}